---
title: "Lab 3 Report"
output: html_document
---

Ben Porter

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

(a)

```{r}
library(ggplot2)
library(GGally)

music <- read.csv("music-plusnew-sub.csv", row.names = 1, stringsAsFactors = FALSE)

music.sub <- subset(music, Type == "Rock" | Type == "Classical")

ggplot(music.sub, aes(x = LAve)) +
  geom_histogram(binwidth = 10) +
  facet_wrap(~ Type, ncol = 1)
```
Classical is centered at a higher value than Rock, as well as Classical is right-skewed and Rock is more left-skewed and Classical has a larger range than Rock.

(b)

```{r}
ggplot(music.sub, aes(x = LVar, y = LAve, color = Type)) +
  geom_point()
```
Rock: large var of amplitude, small average. two groups.

Classical: small var of amplitude, large avg of amplitude.

(c)
```{r}
music.sub2 <- subset(music, Type == "Rock" | Type == "Classical" | Type == "New wave")
ggparcoord(music.sub2, columns = c(4, 3, 5, 6, 7), groupColumn = "Type",
           title = "Parallel Coordinate Plot: Music Types")
```

Rock: large var, max, and ener. low freq.

Classical: large avg and freq. Small var, max and ener.

New wave: Small avg, var and freq.

Rock can be subdivided into 2 types, because there seems to be 2 distinct groups within Rock.

## Exercise 2

(a)
```{r}
bodyfat <- read.csv("bodyfat.csv")

ggpairs(bodyfat[, 1:4])
```

2 outliers.

(b)
```{r}
bf.sub <- subset(bodyfat, Heightinches > 60 & Weightlbs < 300)
ggpairs(bf.sub, columns = 1:4)
```

moderate/strong positive relationship between weight and %bodyfat, and weight and height. No other obvious relationship between other variables.

(c)
```{r}
ggparcoord(bf.sub, 5:14)
```

Two outliers which have large ankle circumference.

(d)
```{r}
bf.sub2 <- subset(bf.sub, AnkleCircm < 33)
ggparcoord(bf.sub2, 5:14)
```

All variables bounce around their centers. No obvious structure.

## Exercise 3

(a)
```{r}
pisamath <- read.csv("pisamathmeans.csv")
summary(pisamath)

pisamath2 <- subset(pisamath, !is.na(acc))
ggpairs(pisamath2, mapping = aes(color = Gender, alpha = 0.5))

```

4978 missing values

(b)
```{r}
ggpairs(pisamath2, mapping = aes(color = Gender, alpha = 0.5))
```

Males and females have similar distributions for all variables. Strong positive linear relationship between all pairs of 7 test scores.

