---
title: 'Homework #2'
author: "Ben Porter"
date: "2/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
```{r}
scores = c(9.3, 8.7, 9.5, 10, 9, 8.9)
scores[-c(which.min(scores), which.max(scores))]
ave = mean(scores[-c(which.min(scores), which.max(scores))])
ave

```

## Question 2
```{r}
ndays = rep(365, times = 21)
ndays[1] = 366
ndays[5] = 366
ndays[9] = 366
ndays[13] = 366
ndays[17] = 366
ndays[21] = 366
ndays
```

## Question 3
```{r}
x = ((-1)^(1:500+1))/1:500
head(x)
sum(x)

```

This series converges and as i increases, each value gets closer and closer to 0, so the bigger the number, the better approximation.

## Question 4
```{r}
Chris_bmi = 70 / 1.8^2
Mary_bmi = 49 / 1.65^2

bmi = cbind(Chris_bmi, Mary_bmi)
bmi

```

## Question 5
```{r}
x = c(1, 1, 4, 6, 4, 7, 5, 9)

#a.
sort(x)

#b.
rank(x, ties.method = "min")

#c.
unique(x)

#d.
y = x[1]
count <- 0
for (i in x) {
if(i < y)  
  count = count+1
y = i
}
print(count)

#e.
b = 5
c = order(abs(x-b))
d = c[1:4]
e = x[c(d)]
mean(e)

#f.
f = sort(abs(x-b))
g = f[f<=2]
h = c[1:length(g)]
i = x[c(d)]
mean(i)
```

## Question 6
```{r}
trunc(.5 + -2:4)

round(.5 + -2:4)
```
The reason that the outputs are different is because trunc always rounds toward 0 and round rounds normally, or to the decimal place specified which is 0 by default, and rounds fives to the even number.\

##Penguins
```{r}
library(palmerpenguins)
```

## Question 1
```{r}
colnames(penguins)
dim(penguins)
summary(penguins)
```

## Question 2
```{r}
head(order(penguins$body_mass_g))
```

## Question 3
```{r}
large = penguins[c(315, 59, 65, 55, 99),]
large$species
```

## Question 4
```{r}
mean(penguins$body_mass_g, na.rm = TRUE)
median(penguins$body_mass_g, na.rm = TRUE)

```

## Question 5
```{r}
#a.
hist(penguins$body_mass_g, main = "Histogram of Body Mass", xlab = "Body Mass", ylab = "Frequency")

#b.
boxplot(penguins$body_mass_g ~ penguins$species, xlab = "Species", ylab = "Body Mass")

#c.
barplot(table(penguins$species), main = "Number of Penguins Per Species", xlab = "Species", ylab = "Number of Penguins")

#d.
plot(penguins$flipper_length_mm, penguins$body_mass_g, col = penguins$sex, main = "Body Mass vs. Flipper Length for Each Sex", xlab = "Flipper Length", ylab = "Body Mass")

```

