---
title: "Lab8"
output:
  html_document:
    df_print: paged
---

Ben Porter

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data on Ames Livability
```{r}
towns <- read.csv("towns2.csv")
dim(towns)
str(towns)
head(towns)
towns$ames <- 0
towns$ames[52] <- 1
```

## Exercise 1
```{r}
library(ggplot2)
library(reshape2)

towns.melt <- melt(towns, id.vars = c(1:3, 14), measure.vars = 4:13)
qplot(value, Score, data = towns.melt, geom = "point") +
  facet_wrap(~ variable) +
  labs(x = NULL)
cor(towns$Score, towns[, 4:13])
```

The variable Climate Diversions has the smallest correlation with livability score, house has a negative correlation and Urban has a low correlation. The rest of the variables have moderate correlations with livability score.

## Exercise 2
```{r}
library(GGally)
ggpairs(towns[, 4:14], upper = list(continuous = wrap("cor", size = 2)),
        lower = list(continuous = wrap("points", stroke = 0, size = .5)))
sapply(towns[, 3:13], sd)
qplot(variable, value, data = towns.melt, geom = "boxplot") +
  labs(x = NULL, y = NULL)
```

The variation for each variable is quite different so it is necessary to use the correlation matrix, or standardize the data, before doing a principal component analysis to account for this.

## Exercise 3
```{r}
towns.pca <- prcomp(towns[, 3:13], scale. = T)  # include Population
summary(towns.pca)
round(towns.pca$rotation, 4)
qplot(1:11, towns.pca$sdev ^ 2, geom = c("point", "line"),
      xlab = "No. of PCs", ylab = "Component Variance (eigenvalue)", main = "Scree Plot")
```

a.

You would need to use 6 PCs in order to explain 80% of the total variation.

b.

```{r}
cumsum(towns.pca$sdev ^ 2)/11

```


## Exercise 4

a.

PC1: The average of Climate, House, Urban, minus the average of Diversion, Economic, Education, Community, Health, Safety, Transportation.

PC2: The average of Population, Climate, Education, Safety, and Urban, minus the average of Economic, Health, and House.

b.

If you want to explain 67% of the variation, You would use 4 PCs because the Cumulative Proportion at PC4 is 67.09% as well as the elbow in the scree plot being at PC4 so using 4 PCs describes a lot of variability while keeping the total number of PCs low.

## Exercise 5
```{r}
head(towns$City[order(towns$Score, decreasing = T)])
head(towns$City[order(towns.pca$x[, 1], decreasing = T)])
qplot(towns$Score, -towns.pca$x[, 1], col = as.factor(towns$ames), asp = 1) +
  labs(x = "Score", y = "negative PC1 score", color = "Ames")
cor(towns$Score, -towns.pca$x[, 1])
```

Mount.Vernon.WA would be rated first using the score variable.

Albemarle.NC would be rated first using the first PC.

While the 2 methods give different cities for the top 6 cities, they do give somewhat similar ratings with Score and PC1 having a moderate correlation of 0.55311.