---
title: "Assignment 3"
author: "Ben Porter"
date: "10/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Assignment 3

###1A.
```{r}
library(car)
library(biotools)
library(GGally)
library(ggplot2)
library(DescTools)


rabbits2 <- read.csv("rabbits2.csv")
head(rabbits2)


rabbits2a = rabbits2
rabbits = rabbits2a[2:7]
rabbits2a$insulin <- as.factor(rabbits2a$insulin)

rabbitsgroup1 = rabbits2a[rabbits2a[,2]=="1", -2]
rabbitsgroup2 = rabbits2a[rabbits2a[,2]=="2", -2]

rabbitsgroup1a = rabbitsgroup1[, -1]
rabbitsgroup2a = rabbitsgroup2[, -1]

group1profile = c(1, colMeans(rabbitsgroup1a))
group2profile = c(2, colMeans(rabbitsgroup2a))

rabbits.data = data.frame(rbind(group1profile, group2profile))
colnames(rabbits.data) = c("insulin", "X1", "X2", "X3", "X4", "X5")
rabbits.data$insulin <- as.factor(rabbits.data$insulin)

ggparcoord(rabbits.data, columns=2:6, groupColumn="insulin", 
             title="")





```

They do not look to be the same for each of the manufacturers. 

###1B.
```{r}



T2result = HotellingsT2Test(rabbitsgroup1a,rabbitsgroup2a)
T2result

T2 = (0.39966*(5*(53 - 1)))/(53-5)
T2
```

Scaled T2 = 0.39966

T2 = 2.164825

Degrees of Freedom of 5 and 48

p-value = 0.8466 > 0.05, cannot reject H0. The mean vectors(profiles) of the 2 manufacturers are equal to each other.


###1C.
```{r}
boxM(rabbits[2:6], rabbits$insulin)


xbar1 <- colMeans(rabbitsgroup1a)

xvar1 <- var(rabbitsgroup1a)

xbar2 <- colMeans(rabbitsgroup2a)

xvar2 <- var(rabbitsgroup2a)

# Compute sample sizes and number of variables
n1 <- nrow(rabbitsgroup1a)
n2 <- nrow(rabbitsgroup2a)
p <- ncol(rabbitsgroup1a)

# Enter the confidence level
level <- 0.95

# Compute degrees of freedom and the multipliers
df1 <- p
df2 <- n1 + n2 - p - 1
df3 <- n1 + n2 - 2
c_T2 <- sqrt((n1 + n2 - 2) * p * qf(level, df1, df2) / (n1 + n2 - p - 1))
levelt <- 1 - (1 - level) / 2
levelb <- 1 - (1 - level) / (2 * p)
c_bon <- qt(levelb, df3)

# Compute pooled covariance matrix 
vpool <- ((n1 - 1) * xvar1 + (n2 - 1) * xvar2) / (n1 + n2 - 2)

# Compute Bonferroni limits
lower_limit <- (xbar1 - xbar2) - c_bon * sqrt(diag(vpool) * ((1 / n1) + (1 / n2)))
upper_limit <- (xbar1 - xbar2) + c_bon * sqrt(diag(vpool) * ((1 / n1) + (1 / n2)))
rbind(lower_limit, upper_limit)
```

p-value (0.09674) > 0.05, so use pooled covariance.

There is no significant difference between the 2 manufacturers for any of X1, X2, X3, X4, and X5.

###1D.
```{r}
library(mvShapiroTest)
mvShapiro.Test(as.matrix(rabbits2[3:7]))

```

p-value: 3.047e-07 < 0.05, so we reject H0. Data are not from 5d normal distribution.


###2A.

```{r}
crudeoil <- read.csv("crudeoil.csv")

crudeoil1 = crudeoil
crudeoil1$Zone <- as.factor(crudeoil1$Zone)

fit.lm <- lm(cbind(X1, X2, X3, X4, X5) ~ Zone, data = crudeoil1)

fit.manova <- Manova(fit.lm)
summary(fit.manova)

```

wilks: 0.116048

F: 18.96779

df: 10 and 98

p-value: 2.22e-16

p-value < 0.05, reject H0. At least two groups have different population mean on 1 of 5 variables.


###2B.
```{r}
boxM(crudeoil1[, -1], crudeoil1$Zone)

# Compute sample sizes and number of variables
n1 <- nrow(crudeoil1[crudeoil1$Zone == 1, -1])
n2 <- nrow(crudeoil1[crudeoil1$Zone == 2, -1])
n3 <- nrow(crudeoil1[crudeoil1$Zone == 3, -1])
p <- ncol(crudeoil1[, -1])
g <- 3

# Enter the confidence level
level <- 0.95

# Compute Bonferroni Confidence Intervals
m <- p * g * (g - 1) / 2  # p dimensions, g (g-1) / 2 combinations (choose 2 out of g)
levelb <- 1 - ((1 - level) / (2 * m))
df <- n1 + n2 + n3 - g
c_bon <- qt(levelb, df)

# Compute Bonferroni limits
xbar1 <- colMeans(crudeoil1[crudeoil1$Zone == 1, -1])
xbar2 <- colMeans(crudeoil1[crudeoil1$Zone == 2, -1])
xbar3 <- colMeans(crudeoil1[crudeoil1$Zone == 3, -1])

var1 <- var(crudeoil1[crudeoil1$Zone == 1, -1])
var2 <- var(crudeoil1[crudeoil1$Zone == 2, -1])
var3 <- var(crudeoil1[crudeoil1$Zone == 3, -1])

lower_limit_1v2_unpool <- (xbar1 - xbar2) - c_bon * sqrt(diag(var1) / n1 + diag(var2) / n2)
upper_limit_1v2_unpool <- (xbar1 - xbar2) + c_bon * sqrt(diag(var1) / n1 + diag(var2) / n2)

lower_limit_1v3_unpool <- (xbar1 - xbar3) - c_bon * sqrt(diag(var1) / n1 + diag(var3) / n3)
upper_limit_1v3_unpool <- (xbar1 - xbar3) + c_bon * sqrt(diag(var1) / n1 + diag(var3) / n3)

lower_limit_2v3_unpool <- (xbar2 - xbar3) - c_bon * sqrt(diag(var2) / n2 + diag(var3) / n3)
upper_limit_2v3_unpool <- (xbar2 - xbar3) + c_bon * sqrt(diag(var2) / n2 + diag(var3) / n3)

rbind(lower_limit_1v2_unpool, upper_limit_1v2_unpool,
      lower_limit_1v3_unpool, upper_limit_1v3_unpool,
      lower_limit_2v3_unpool, upper_limit_2v3_unpool)




```

p-value (0.01542) < 0.05, so use individual covariance.

There is no difference in mean vanadium, iron, beryllium, and saturated hydrocarbons levels for crude oil extracted from the Wilhelm and Sub-Mulinia zones.

Crude oil extracted from the Wilhelm zone has a significantly higher mean level of aromatic hydrocarbons than crude oil extracted at either the Sub-Milina or the Upper zones. 

Crude oil extracted from the Wilhelm zone has a significantly lower mean level of vanadium and beryllium than crude oil extracted at the Upper zones.

Crude oil extracted from the Wilhelm zone has a significantly higher mean level of iron and saturated hydrocarbons than crude oil extracted at the Upper zones. 

There is no difference in mean iron and aromatic hydrocarbons levels for crude oil extracted from the Sub-Mulinia and Upper zones.

Crude oil extracted from the Sub-Mulinia zone has a significantly lower mean level of vanadium and beryllium than crude oil extracted at the Upper zones.

Crude oil extracted from the Sub-Mulinia zone has a significantly higher mean level of saturated hydrocarbons than crude oil extracted at the Upper zones. 






