---
title: 'Homework #4'
author: "Ben Porter"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Question 1
# a.
```{r}
A <- diag(1, 4)
A
B <- diag(c(3.1, 2.9, 1.2), 3)
B
C <- matrix(B[1,1], nrow = 1, ncol = 1)
C
D <- matrix(1:3, nrow = 3, ncol = 1)
D
E <- matrix(1:10, nrow = 2, ncol = 5, byrow = TRUE)
E
F <- matrix(-3:2, nrow = 3, ncol = 2)
F


```

# b.
```{r}
#i.
A[1, , drop = FALSE]

#ii.
A[1, ]

#iii.
B[2, 2] <- 0
B

#iv.
t(A) + 1:4

#v.
G <- cbind(E, t(F))
G

#vi.
H <- rbind(t(E), F)
H

```

# c.
```{r}
#i.
F %*% E

#ii.
c(D)*c(t(D))

#iii.
5*B + D%*%t(D)

#iv.
B + c(D)

#v.
diag(B)

#vi.
solve(t(F)%*%F) %*% E

```

## Question 2
# a.
```{r}
#a.
library(datasets)
data <- data.frame(iris)
dat <- subset(data, data$Species %in% 'setosa' | data$Species %in% 'versicolor')

```

# b.
```{r}
dat$Y <- matrix(ifelse(dat$Species %in% 'setosa', 1, 0))
dat$Y

```

# c.
```{r}
X <- as.matrix(dat[, c('Sepal.Length', 'Sepal.Width', 'Petal.Length', 'Petal.Width')])
intercept <- 1
X <- cbind(intercept, X)
head(X)
```

# e.
```{r}
average <- ((X[,'Sepal.Length'] + X[,'Sepal.Width'])/2)
X <- cbind(X, average)
head(X)
```

# g.
```{r}
#g.
 #betaHat <- solve(t(X) %*% X) %*% t(X) %*% dat$Y

```
We cannot calculate the least squares estimate because you can't invert t(X) %*% X after adding the average column because it is colinear with Sepal.Length and Sepal.Width, the columns that are directly used to calculate it.

# h.
```{r}
I <- diag(1, ncol(X))
Lambda <- 1
betaHat <- solve(t(X) %*% X + Lambda * I) %*% t(X) %*% dat$Y
head(betaHat)
```

# i.
```{r}
YHat <- X %*% betaHat
head(YHat)
RSS <- t(dat$Y - YHat) %*% (dat$Y - YHat)
RSS
list1 <- list(betaHat = betaHat, YHat = YHat, RSS = RSS)
head(list1)
```

#j.
```{r}
Lambda <- .5
betaHat <- solve(t(X) %*% X + Lambda * I) %*% t(X) %*% dat$Y
YHat <- X %*% betaHat
RSS <- t(dat$Y - YHat) %*% (dat$Y - YHat)
RSS

Lambda <- .3
betaHat <- solve(t(X) %*% X + Lambda * I) %*% t(X) %*% dat$Y
YHat <- X %*% betaHat
RSS <- t(dat$Y - YHat) %*% (dat$Y - YHat)
RSS

Lambda <- .1
betaHat <- solve(t(X) %*% X + Lambda * I) %*% t(X) %*% dat$Y
YHat <- X %*% betaHat
RSS <- t(dat$Y - YHat) %*% (dat$Y - YHat)
RSS

```

A value of .1 gives me seems to give me the smallest RSS out of the values I tested(.5, .3, .1).

# k.
```{r}
Lambda <- .1
betaHat <- solve(t(X) %*% X + Lambda * I) %*% t(X) %*% dat$Y
YHat <- X %*% betaHat
RSS <- t(dat$Y - YHat) %*% (dat$Y - YHat)

YHat <- ifelse(YHat < 0.5, 0, 1)
miscalculated <- YHat != dat$Y
length(miscalculated[miscalculated == TRUE])
```

## Question 3
# a.
```{r}
string <- c("string")
vector <- c(1:5)
logical <- c(TRUE, FALSE)
list <- list(1, "two", 3)
listofall <- list(string, vector, logical, list)
listofall
```

#b.
```{r}
myList <- vector(mode = "list", length = 10)
myList[1:10] <- 1:10
str(myList)



```

## Question 4
# a.
```{r}
l <- list(A = c("Red", "Green", "Black"), 
          B = matrix(1:6, nrow = 2), 
          C = list("Python", "PHP", "Java"))
length(l)
```

# b.
```{r}
thirdentry <- l[[3]]
thirdentry
```

# c.
```{r}
thirdnested <- list(thirdentry[3])
thirdnested
```

# d.
```{r}
l[[2]][2,]
```

# e.
```{r}
str(l[[2]]+1)
```

# f.
```{r}
l <- l[-1]
l


```

# g.
```{r}
D <- data
l1 <- append(l, D)
str(l1)
```

## Question 5
# a.
```{r}
library(lattice)
barley1 <- barley
dim(barley1)
```

# b.
```{r}
ncol(barley1)

```

# c.
```{r}
barley1 <- barley1[-1,]
dim(barley1)
```

# d.
```{r}
barley1 = rbind(barley1, c(50, "Svansota", 1931, "Duluth"))
```

# e.
```{r}
head(barley1, 3)
tail(barley1, 6)
```

# f.
```{r}
barley1["year"] <- as.numeric(as.character(unlist(barley1["year"])))
str(barley1)
```

# g.
```{r}
barley1["newYield"] <- barley$yield/0.4047
str(barley1)
```

# h.
```{r}
barley1 <- subset(barley1,select = -yield)
str(barley1)

```

# i.
```{r}
barley2 <- subset(barley1,select = -year)
barley2 <- barley2[grepl("^No.", barley2$variety),]
str(barley2)

```



