---
title: "Lab 4 Report"
author: "Ben Porter"
date: "9/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Exercise 1
```{r}
sweat <- read.table("sweat.dat", header = F, col.names = c("subject", "x1", "x2", "x3"))

library(ggplot2)
library(GGally)
ggpairs(sweat[, 2:4])

xbar <- sapply(sweat[, 2:4], mean)  # same as colMeans function
xbar
scor <- cor(sweat[, 2:4])
scor
svar <- var(sweat[, 2:4])           # same as cov function
svar
det(svar)                           # determinant of covariance matrix
sum(diag(svar))                     # trace of covariance matrix
sum(diag(scor))
```

(a)

```{r}
library(ggplot2)
library(GGally)

sweat <- read.table("sweat.dat", header = F, col.names = c("subject", "x1", "x2", "x3"))
ggpairs(sweat[, 2:4])

```
They have somewhat linear relationships with eachother, although x2 and x3 do not have a clear relationship.
(b)
no outliers. elliptical
(c)
bell shape. (small skewness)
(d)
```{r}

xbar <- sapply(sweat[, 2:4], mean)
xbar
scor <- cor(sweat[, 2:4])
scor
svar <- var(sweat[, 2:4]) 
svar
det(svar)
sum(diag(svar))
sum(diag(scor))

```
The trace of the correlation matrix is the number of variables because the daigonal of the correlation matrix is 1 for each variable.


## Exercise 2

```{r}
sweat_qqplot <- data.frame(x = c(sweat[, 2], sweat[, 3], sweat[, 4]),
                           variable = rep(colnames(sweat[, 2:4]), each = nrow(sweat)))
ggplot(sweat_qqplot, aes(sample = x)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ variable, scales = "free", nrow = 2)

sapply(sweat[, 2:4], shapiro.test)


```
All 3 columns look normally distributed based on the qq plot and all 3 p-values > 0.05.

## Exercise 3

(a)
H0: are from 3-dim multivariate distribution.
Ha: are not from 3-dim multivariate distribution.


(b)
```{r}
library(mvShapiroTest)
mvShapiro.Test(as.matrix(sweat[, 2:4]))
```
cannot reject H0, because p-value (0.2567) > 0.05.


## Exercise 4
(a)
H0: mu = (4, 50, 10)
Ha: mu not = (4, 50, 10)


(b)
Normality: are from 3-dim multivariate normal distribution.
Independence: are sampled independently.


(c)
```{r}
library(DescTools)
t2test <- HotellingsT2Test(sweat[, 2:4], mu = c(4, 50, 10))  # T.2 is the scaled value of the T2 statistic
t2test
n <- nrow(sweat[, 2:4])
p <- ncol(sweat[, 2:4])
t2test$statistic * p * (n - 1) / (n - p)


```
T2: 9.738. F: 2.9045. D.f: 3 and 17. P-value: 0.06493.


(d)

Cannot reject H0, because p-value (0.06493) > 0.05.

## Exercise 5
```{r}
boards <- read.csv("board.stiffness.csv")
head(boards)
summary(boards[, 2:5])
ggpairs(boards[, 2:5])

newboards <- read.csv("newboard.stiffness.csv")
newboards

allboards <- rbind(boards, newboards)
dim(allboards)
allboards$wood <- "original"
allboards$wood[allboards$board %in% 1001:1003] <- "new"
allboards$wood <- factor(allboards$wood)
allboards
ggpairs(allboards[, 2:6], mapping = aes(color = wood, alpha = 0.5))

newb <- as.matrix(newboards[, 2:5])
bmean <- colMeans(boards[, 2:5])
bmean <- as.matrix(bmean)
bvar <- var(boards[, 2:5])
n <- dim(boards)[1]
p <- length(bmean)
one <- matrix(c(1, 1, 1), ncol = 1)
diff <- newb - one %*% t(bmean)
sdiff <- ((n + 1) / n) * bvar
d22 <- diff %*% solve(sdiff) %*% t(diff)
d2 <- diag(d22)

prob <- 1 - pchisq(d2, p)
prob
```

1st (4.77e-04) and 3rd (3.37e-09): p-value < 0.05. They are not from the same type of wood.

2nd (0.9998): p-value > 0.05. It is from the same type of wood.


## Exercise 6
```{r}
library(mvtnorm)

set.seed(40713)  # Do this before your first random generator!!!!

xsigma <- matrix(c(1, -0.5, -0.5, 1), ncol = 2, byrow = T)

# Simulate data from multivariate normal distribution
sim.mvn <- data.frame(rmvnorm(100, mean = c(0, 0), sigma = xsigma))
head(sim.mvn)
qplot(X1, X2, data = sim.mvn) + theme(aspect.ratio = 1)
mvShapiro.Test(as.matrix(sim.mvn))

# Simulate data from multivariate t distribution
sim.mvt <- data.frame(rmvt(100, sigma = xsigma))
head(sim.mvt)
qplot(X1, X2, data = sim.mvt) + theme(aspect.ratio = 1)
mvShapiro.Test(as.matrix(sim.mvt))
```
Yes, this matched my expectations because the shapiro-wilk test for mvn had a p-value (0.5789) > 0.05 so it doesn't reject H0(is from mvn distribution) and the test for the mvt had a p-value (2.2e-16) < 0.05 so it does reject H0(is from mvn distribution).
