---
title: "lab 2 Report"
output: html_document
---
Benjamin Porter


```{r}
tips <- read.csv("http://www.ggobi.org/book/data/tips.csv")
Ames <- read.csv("Ameshousing.csv", stringsAsFactors = T)
head(Ames)
dim(Ames)
```

## Exercise 1
(a)
```{r}
dim(Ames)
```
168

(b)
```{r}
names(Ames)
```

## Exercise 2
(a)
```{r}
LivingArea_median = median(Ames$LivingArea)
LivingArea_median
LivingArea_mean = mean(Ames$LivingArea)
LivingArea_mean
LivingArea_std = sd(Ames$LivingArea)
LivingArea_std
n = length(Ames$LivingArea)
n
stderr_LivingArea = LivingArea_std / sqrt(n)
stderr_LivingArea


```

(b)
```{r}
Ames$LivingArea
tapply(Ames$LivingArea, Ames$Neighborhood, median)
tapply(Ames$LivingArea, Ames$Neighborhood, mean)
LivingArea_sd1 <- tapply(Ames$LivingArea, Ames$Neighborhood, sd)
LivingArea_sd1
samplesizes1 <- tapply(Ames$LivingArea, Ames$Neighborhood, length)
samplesizes1
LivingArea_sderr1 <- LivingArea_sd1 / sqrt(samplesizes1)
LivingArea_sderr1


```

(c)
```{r}
Ames1 = Ames[Ames$LivingArea < 1800, ]
nrow(Ames1)
mean(Ames1$SalePrice)
sd(Ames1$SalePrice) / sqrt(length(Ames1$SalePrice))
```

(d)
```{r}
Ames2 = Ames[Ames$LivingArea < 2000, ]
nrow(Ames2)
mean(Ames2$SalePrice)
sd(Ames2$SalePrice) / sqrt(length(Ames2$SalePrice))
```

(e)
```{r}
tapply(Ames1$SalePrice, Ames1$Neighborhood, median)
tapply(Ames1$SalePrice, Ames1$Neighborhood, mean)
tapply(Ames1$SalePrice, Ames1$Neighborhood, sd) / sqrt(tapply(Ames1$SalePrice, Ames1$Neighborhood, length))

```


## Exercise 3
(a)
```{r}
library(ggplot2)

lm(SalePrice ~ LivingArea, data = Ames)
ggplot(Ames, aes(x = LivingArea, y = SalePrice)) +
  geom_point() +
  geom_smooth() +
  geom_abline(intercept = -21769.1, slope = 159.5, col = "red") +
  labs(x = "Living Area", y = "Sale Price", title = "Sale Price Vs. Living Area")

```
As living area increases, the mean of the sale price increases.

(b)
As the living area increases, the variance of the sale price increases.

## Exercise 4
(a)
```{r}
Ames$pricesqft <- Ames$SalePrice / Ames$LivingArea

ggplot(Ames, aes(x = pricesqft)) +
  geom_histogram() +
  labs(x = "Price per sq. foot of Living Area")
```
bimodal, right-skewed, and centered around 150, majority is 100-200.

(b)
```{r}
ggplot(Ames, aes(x = pricesqft)) +
  geom_histogram(binwidth = 20) +
  labs(x = "Price per sq. foot of Living Area")
```


## Exercise 5
(a)

```{r}
ggplot(Ames, aes(x = LivingArea, y = SalePrice)) +
  geom_point() +
  geom_smooth() +
  facet_grid(. ~ Neighborhood)
```
Positive relationship between Living Area and Sale Price for all 3 neighborhoods.

Strong linearity in CollCr and NridgHt.


(b)

```{r}
ggplot(Ames, aes(x = Neighborhood, y = pricesqft)) +
  geom_boxplot()
```
Glibert lowest center, then Collgcr, NridgHt.
CollgCr has largest IQR, followed by NridgHt, then Gilbert
CollgCr is skewed to the right.


(c)

```{r}
ggplot(Ames, aes(x = Neighborhood, y = LotArea)) +
  geom_boxplot()

```
Glibert lowest center, then Collgcr, NridgHt.
NridgHt has largest IQR, followed by Gilbert, then CollgCr.
Gilbert is skewed to the right.


## Exercise 6
(a)
Positive relationship in all pairs.
Strong correlation between SalePrice and LivingArea, SalePrice and Garage.
Weak correlation between SalePrice and Bedroom, Bedroom and GarageArea.
```{r}
AmesX = Ames[, 1:5]
cor(AmesX)
```


(b)
You can see Bedroom is a discrete variable.
You can see the variance increase on the scatterplots.
You can also see that general shape of the relationship between each of the variables.


```{r}
library(GGally)
ggpairs(AmesX)
```

