---
title: "Homework 2 report"
author: "Ben Porter"
date: "9/29/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 1:

a.
```{r}
library(Rcpp)
library(biotools)
library(ggplot2)
library(GGally)
library(reshape2)
library(mvShapiroTest)
library(DescTools)
cork = read.csv("D:/school/Stat/Stat 475/Homework 2/cork.csv")


cork1 = cork[2:5]
colMeans(cork1)
```

b.
```{r}
boxplot(cork1)
```

The boxplots suggest slight differences. While the boxplots show very similar distributions, X1(North) looks to have a slightly higher average and distribution than the others, as well as X3(South) has the largest range/most variability followed by X4(West), then X1(North), and then X2(East).

c.
```{r}
cor(cork1)

ggpairs(cork1)
```

All correlations are positive and strong. The scatterplot matrix shows a positive linear trend between each of the variables.

d.
```{r}
ggplot(melt(cork, id.vars = 1, measure.vars = 2:5), aes(sample = value)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ variable, scales = "free", nrow = 2)

# Univariate Shapiro-Wilk test
apply(cork1, 2, shapiro.test)

# Multivariate Shapiro-Wilk test
mvShapiro.Test(as.matrix(cork1))
```

X3(South)(p-value 0.007059 < 0.05) rejects H0. X3(South) is not a normal distribution based on the univariate Shapiro-Wilk test. The QQ plot of X3 also supports non-normality.

X1(North)(p-value 0.0179 > 0.05), X2(East)(p-value 0.01351 > 0.05), and X4(West)(p-value 0.1185 > 0.05) cannot reject H0. X1(North), X2(East), and X4(West) are from normal distributions based on the univariate Shapiro-Wilk test. The QQ plots of X1, X2, and X4 also support normality.

The multivariate Shapiro-Wilk test(p-value 0.03314 < 0.05) rejects H0. This means the data is not from a 4d normal distribution.

e.
```{r}
HotellingsT2Test(cork1, mu = c(47, 47, 47, 47), test = "f")

T2 = (6.3892*(4*(28 - 1)))/(28-4)
T2
```

Scaled T2 = 6.3892

T2 = 28.7514

Degrees of Freedom of 4 and 24

p-value = 0.001197 < 0.05, reject H0. The population means of X1, X2, X3, and X4 are not equal to 47.

f.
```{r}
HotellingsT2Test(cork1, test = "f")
T2b = (72.485*(4*(28 - 1)))/(28-4)
T2b
```

Scaled T2 = 72.485

T2 = 326.1825

Degrees of Freedom of 4 and 24

p-value = 4.814e-13 < 0.05, reject H0. The population means of X1, X2, X3, and X4 are not equal to each other.

g.
```{r}
TB.conf.int <- function(X, level = 0.95)
{ 
  # Convert X to a matrix, if it is not a matrix already, from
  # vectors or data frames.
  X <- as.matrix(X)
  
  # Set n to the number of observations, p to the number of variables.
  n <- nrow(X)
  p <- ncol(X)
  
  # Stop if arguments are invalid.
  if (!is.numeric(X))
  {
    stop("Data must be numeric")
  }
  
  if (n < p)
  {
    stop("Must have at least as many observations as variables")
  }
  
  if (!is.numeric(level) || length(level) != 1 || level <= 0 || level >= 1)
  {
    stop("Confidence level must be between 0 and 1")
  }
  
  # Create a matrix A in which each column represents
  # a difference between two pairs of means
  np <- p * (p - 1) / 2
  A <- matrix(c(0), ncol = np, nrow = p)
  nc <- 0
  for (i in 1:(p - 1)) {
    for (j in 1:(p - i)) {
      A[i, nc + j] <- 1
      A[i + j, nc + j] <- -1
    }
    nc <- nc + (p - i)
  }
  
  # Create a matrix that will hold the confidence intervals.
  CI <- matrix(NA, 2, ncol(A))
  rownames(CI) <- c("lower", "upper")
  colnames(CI) <- colnames(A)
  
  CIB <- matrix(NA, 2, ncol(A))
  rownames(CIB) <- c("lower", "upper")
  colnames(CIB) <- colnames(A)
  
  CIT <- matrix(NA, 2, ncol(A))
  rownames(CIT) <- c("lower", "upper")
  colnames(CIT) <- colnames(A)
  
  # Find F distribution quantile for T-squared confidence intervals.
  F <- qf(level, p, n - p)
  
  # Find t distribution percentile for Bonferroni confidence intervals
  alpha <- (1 - level) / 2 / ncol(A)
  levelB <- 1 - alpha 
  tB <- qt(levelB, n - 1)
  t <- qt(1 - (1 - level) / 2, n - 1)
  
  # Compute the sample covariance matrix of the original variables.
  C <- cov(X)
  
  # Find the confidence intervals for the specified linear combinations.
  for (i in 1:ncol(A))
  { 
    # Find the sample mean and variance of this linear combination.
    m <- mean(X %*% A[, i])
    v <- t(A[, i]) %*% C %*% A[, i]
    
    # Find the confidence interval for this difference.
    CI[1, i] <- m - sqrt((p * (n - 1) / n / (n - p)) * F * v)
    CI[2, i] <- m + sqrt((p * (n - 1) / n / (n - p)) * F * v)
    
    CIB[1, i] <- m - tB * sqrt(v / n)
    CIB[2, i] <- m + tB * sqrt(v / n)
    
    CIT[1, i] <- m - t * sqrt(v / n)
    CIT[2, i] <- m + t * sqrt(v / n)
  }
  
  # Print the confidence intervals.
  cat(" T-squared simultaneous confidence intervals: \n\n")
  print(CI)
  
  cat("\n\n Bonferroni simultaneous confidence intervals: \n\n")
  print(CIB)
  
  cat("\n\n One-at-a-Time t confidence intervals: \n\n")
  print(CIT)
}

TB.conf.int(cork1)

```

According to the Bonferroni simultaneous confidence intervals:

μ1 > μ2

μ1 > μ4

μ3 > μ4

No significant difference between μ1 and μ3, μ2 and μ3, and μ2 and μ4.

On average, the North side(X1) has thicker cork deposits than both the East side(X2) and the West side(X4). Also, the South side(X3) has thicker cork on average than the West side(X4).

Question 2:

a.
```{r}
bnotes = read.csv("D:/school/Stat/Stat 475/Homework 2/bnotes.csv")
bnotes1 = bnotes[2:7]

# Univariate Shapiro-Wilk test
apply(bnotes1, 2, shapiro.test)
```

Value of W:

X1 = 0.98374

X2 = 0.96431

X3 = 0.96637

X4 = 0.97338

X5 = 0.97456

X6 = 0.95878

p-value:

X1 = 0.2567

X2 = 0.008254

X3 = 0.01174

X4 = 0.04035

X5 = 0.04984

X6 = 0.003295

X2(p-value 0.008254 < 0.05), X3(p-value 0.01174 < 0.05), X4(p-value 0.04035 < 0.05), X5(p-value 0.04984 < 0.05), and X6(p-value 0.003295 < 0.05) rejects H0. X2, X3, X4, X5, and X6 are not normal distributions based on the univariate Shapiro-Wilk test.

X1(p-value 0.2567 > 0.05) cannot reject H0. X1 is from a normal distributions based on the univariate Shapiro-Wilk test.

All of the QQ plots are relatively straight lines, implying normality based on these plots.

b.
```{r}
# Multivariate Shapiro-Wilk test
mvShapiro.Test(as.matrix(bnotes1))
```

The Multivariate Shapiro-Wilk test(p-value 0.02399 < 0.05) rejects H0. This means the data is not from a 6-dimensional normally distributed model according to this test.

c.
```{r}
S = cov(bnotes1)
Sinverse = solve(S)

xbar = as.matrix(colMeans(bnotes1))
x1 = c(214.9, 130.5, 130.2, 8.4, 11.6, 138.4)
diff1 = x1 - xbar
dsquared1 = (100/101)*t(diff1) %*% Sinverse %*% diff1
dsquared1
dchisq(dsquared1, 6)

x2 = c(215.9, 129.5, 130.6, 7.9, 12.1, 140.8)
diff2 = x2 - xbar
dsquared2 = (100/101)*t(diff2) %*% Sinverse %*% diff2
dsquared2
dchisq(dsquared2, 6)


x3 = c(215.3, 130.7, 130.6, 8.1, 11.7, 142.4)
diff3 = x3 - xbar
dsquared3 = (100/101)*t(diff3) %*% Sinverse %*% diff3
dsquared3
dchisq(dsquared3, 6)


x4 = c(214.9, 130.1, 129.8, 8.7, 10.9, 141.8)
diff4 = x4 - xbar
dsquared4 = (100/101)*t(diff4) %*% Sinverse %*% diff4
dsquared4
dchisq(dsquared4, 6)


x5 = c(215.9, 131.3, 129.2, 8.5, 11.5, 138.4)
diff5 = x5 - xbar
dsquared5 = (100/101)*t(diff5) %*% Sinverse %*% diff5
dsquared5
dchisq(dsquared5, 6)
```

Bank note #1:

D squared = 49.43481

P(D squared = 49.43481) = 2.813931e-09

Bank note #2:

D squared = 43.09305

P(D squared = 43.09305) = 5.095149e-08

Bank note #3:

D squared = 21.44749

P(D squared = 21.44749) = 0.0006329496

Bank note #4:

D squared = 7.322699

P(D squared = 7.322699) = 0.08612287

Bank note #5:

D squared = 86.53376

P(D squared = 86.53376) = 7.580258e-17

I would determine that bank notes 1, 2, 3, and 5 are forgeries, as their p-values are much too small to be genuine.

Bank note 4 could possibly be a genuine bank note. Bank note 4 has a p-value of 0.08612287 which is possible with a sample size of 100 we would expect(100 * 0.08612287) 8.61 notes with this d squared.





