---
title: "Lab9"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ben Porter

## Life Expectancy Data
```{r}
library(ggplot2)
library(GGally)

life <- read.csv("life.csv", row.names = 1)
head(life)
lifestd <- scale(life, center = T, scale = T)
ggpairs(as.data.frame(lifestd), lower = list(continuous = "smooth"))
lifecor <- cor(life)
lifecor
```

## Use Principal Component Estimation for Factor Loadings
```{r}
# First do a principal component analysis
lifepc <- prcomp(life, scale = T, center = T)
summary(lifepc)

# Scree plot for principal components
qplot(1:ncol(life), lifepc$sdev ^ 2, geom = "line",
      xlab = "Component Number", ylab = "Component Variance (eigenvalue)", main = "Scree Plot")

# Display coefficients for principal components (eigenvectors)
lifepc$rotation

# Compute factor loadings for factors using principal components from the sample correlation matrix
lifepc$loadings <- lifepc$rotation %*% diag(lifepc$sdev)
lifepc$loadings

# 1. Apply the varimax rotation to the 3-factor model
varimax(lifepc$loadings[, 1:3])

# 2. Apply the varimax rotation to the 2-factor model
lpcv <- varimax(lifepc$loadings[, 1:2])
print(lpcv$loadings, cutoff = 0)
# Compute communalities and uniquenesses
lpcv.com <- diag(lpcv$loadings %*% t(lpcv$loadings))
lpcv.uniq <- diag(lifecor) - lpcv.com
lpcvtable <- cbind(lpcv$loadings, lpcv.com, lpcv.uniq)
colnames(lpcvtable) <- c("Factor1", "Factor2", "Communality", "Uniqueness")
lpcvtable
# Compute factor scores for the 2-factor model
lpcvscore <- lifestd %*% solve(lpcv$loadings %*% t(lpcv$loadings) + diag(lpcv.uniq)) %*% lpcv$loadings
lpcvscore
qplot(lpcvscore[, 1], lpcvscore[, 2], label = abbreviate(rownames(life), 9), size = I(2.5), geom = "text",
      xlab = "Factor 1", ylab = "Factor 2", main = "Varimax Rotated Factors")
```

## Maximum likelihood estimation of factor loadings
```{r}
# Compute p-values for tests for numbers of factors
sapply(1:3, function(f) factanal(life, factors = f, method = "mle")$PVAL)

# 1. Compute loadings for a 3-factor model with no rotation. Factor scores are not computed.
factanal(life, factors = 3, method = "mle", rotation = "none")

# 2. Apply the varimax rotation (default) to the factor loadings for the 3-factor model.  
life_varimax <- factanal(life, factors = 3, method = "mle", rotation = "varimax", scores = "regression")
life_varimax
# Print all of the loading values
print(loadings(life_varimax), cutoff = 1e-05)
# Plot factor scores
scores <- life_varimax$scores
scores
ggpairs(as.data.frame(scores), mapping = aes(label = abbreviate(rownames(life), 9)),
        lower = list(continuous = function(data, mapping) ggplot(data, mapping) + geom_text(size = 2)))

# 3. Compute varimax rotation for the 2-factor model
life_varimax2 <- factanal(life, factors = 2, method = "mle", rotation = "varimax", scores = "regression")
life_varimax2
```

## Pain Data

## Exercise 1
(a)
```{r}
paincor <- matrix(c(1.0, -0.04, 0.61, 0.45, 0.03, -0.29, -0.30, 0.45, 0.30,
                    -0.04, 1.00, -0.07, -0.12, 0.49, 0.43, 0.30, -0.31, -0.17,
                    0.61, -0.07, 1.00, 0.59, 0.03, -0.13, -0.24, 0.59, 0.32,
                    0.45, -0.12, 0.59, 1.00, -0.08, -0.21, -0.19, 0.63, 0.37,
                    0.03, 0.49, 0.03, -0.08, 1.00, 0.47, 0.41, -0.14, -0.24,
                    -0.29, 0.43, -0.13, -0.21, 0.47, 1.00, 0.63, -0.13, -0.15,
                    -0.30, 0.30, -0.24, -0.19, 0.41, 0.63, 1.00, -0.26, -0.29,
                    0.45, -0.31, 0.59, 0.63, -0.14, -0.13, -0.26, 1.00, 0.40,
                    0.30, -0.17, 0.32, 0.37, -0.24, -0.15, -0.29, 0.40, 1.00),
                  ncol = 9, byrow = T)
paincor


painpc <- princomp(covmat = paincor, cor = T)
qplot(1:ncol(paincor), painpc$sdev ^ 2, geom = "line",
      xlab = "Component Number", ylab = "Component Variance (eigenvalue)", main = "Scree Plot")
```

3 PCs because the scree plot starts to flatten out after 3 PCs.

(b)
```{r}
summary(painpc)
```

The first PC explains 0.3824117 of the total variation while PC2 explains 0.2144174 of the variation. 5 PCs are needed to explain at least 80% of the total variance.

(c)
```{r}
painpc$loadingf <- painpc$loadings %*% diag(painpc$sdev)
painpc$loadingf

varpcf <- colSums(painpc$loadingf ^ 2)
varpercent <- varpcf / sum(diag(paincor))
cumpercent <- cumsum(varpercent)
vartable <- rbind(varpcf, varpercent, cumpercent)
rownames(vartable) <- c("SS loadings", "Proportion Var", "Cumulative Proportion")
vartable
```

Factor 1: A contrast between assigning responsibility to the medical profession versus accepting personal responsibility for controlling pain.

Factor 2: An average of the first 8 questions.

0.5968291 of the total variance.

(d)
```{r}
painpc.com <- diag(painpc$loadingf[, 1:2] %*% t(painpc$loadingf[, 1:2]))
painpc.uniq <- diag(paincor) - painpc.com

painpctable <- cbind(painpc$loadingf[, 1:2], painpc.com, painpc.uniq)
colnames(painpctable) <- c("Factor1", "Factor2", "Communality", "Uniqueness")
painpctable
```

Each column of factor loading matrix is the eigen vector times the square root of corresponding eigen-value.

(e)
```{r}
painpc.2f <- varimax(painpc$loadingf[, 1:2])
painpc.2f$loadings
```

- F1: an external responsibility for managing pain.

- F2: a personal responsibility for managing pain.

The rotated F1 is the external responsibility and the rotated F2 is the personal responsibility while the un-rotated F1 is the difference between the external and personal responsibility and the un-rotated F2 is the average.

(f)
```{r}
painpc.4f <- varimax(painpc$loadingf[, 1:4])
painpc.4f$loadings
```

The rotation with 4 factors is better because it explains more variability than the rotation with the 4 factors.

(g)

- F1: an external responsibility for managing pain.

- F2: a personal (internal) responsibility for managing pain related to individuals actions.

- F3: a personal (internal) responsibility for managing pain and not being careless.

- F4: avoiding pain is a matter of luck.

## Exercise 2
(a)
```{r}
mlefact2 <- factanal(covmat = paincor, cor = T, factors = 2, n.obs = 123, method = "mle", rotation = "varimax")
mlefact2
```

Factor1: 27.7\%, Factor2: 21.2\%.

(b)
```{r}
summary(painpc)
```

First 2 factors: 48.9\%. First 2 PC: 59.6\%.

(c)
```{r}
mlefact2$loadings
mlefact2$uniquenesses
mlefact2.com <- diag(mlefact2$correlation) - mlefact2$uniquenesses
mlefact2.com
factable <- cbind(mlefact2$loadings, mlefact2.com, mlefact2$uniquenesses)
colnames(factable) <- c("Factor1", "Factor2", "Communality", "Uniqueness")
factable
```

3rd variable has the largest proportion (68.2\%) of variation explained by 2-factor model. 9th variable has the least proportion (23.4\%) of variation. 

(d)

- F1: an external responsibility for managing pain.

- F2: a personal responsibility for managing pain.

(e)
```{r}
painpctable
```

The principal component method explains a larger proportion of the variances then the MLE.

(f)
```{r}
sapply(1:5, function(f) factanal(covmat = paincor, cor = T, n.obs = 123, factors = f, method = "mle")$PVAL)
```

We need 3 factors based on the likelihood ratio tests because 3 factors are above 0.05.

```{r}
# Compute the residual matrix
mlefact2.res <- mlefact2$correlation - (mlefact2$loadings %*% t(mlefact2$loadings) + diag(mlefact2$uniquenesses))
mlefact2.res
```

